---
title: "PA1_template"
author: "Miguel Adarlo"
date: "2023-07-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reproducible Research - Course Project 1

## Loading and Preprocessing the Data
``` {r loading}
unzip(zipfile = "activity.zip")
activity <- read.csv("activity.csv")
```
## What is mean total number of steps taken per day?
``` {r hist}
library(dplyr)
total_steps <- activity %>% group_by(date) %>% summarize(daily_steps = sum(steps, na.rm = TRUE))

hist(total_steps$daily_steps, main = "Histogram of Daily Steps", xlab="Total Number of Steps Taken Each Day")
```
Calculate the mean:
``` {r meancalc}
mean(total_steps$daily_steps, na.rm = TRUE)
```
Calculate the median:
``` {r mediancalc}
median(total_steps$daily_steps, na.rm = TRUE)
```
## What is the average daily activity pattern?
```{r avgdaily}
steps_interval <- activity %>% group_by(interval) %>% summarize(interval_steps = mean(steps, na.rm = TRUE))

plot(steps_interval$interval_steps ~ steps_interval$interval, type="l", xlab = "Time Interval", ylab = "Average Number of Steps", main = "Steps per 5 Minute Interval")
```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
``` {r max}
max_interval <- steps_interval$interval[which.max(steps_interval$interval_steps)]
max_steps <- max(steps_interval$interval_steps)

paste("Max interval:", max_interval," Number of steps:", max_steps)
```
## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
``` {r NAcalc}
sum(is.na(activity$steps))
```
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Strategy used was to fill in the mean steps for that day for each 5 minute interval, as calculated previously.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

``` {r impute}
activity_impute <- activity  
for (i in 1:nrow(activity)){
        if(is.na(activity$steps[i])){ #Checks for NA.
                activity_impute$steps[i] <- steps_interval$interval_steps[activity_impute$interval[i] == steps_interval$interval]
        }
}
```
This code made use of activity and steps_interval to form the new dataset.

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

``` {r newhist}
total_steps <- activity_impute %>% group_by(date) %>% summarize(daily_steps = sum(steps, na.rm = TRUE))

hist(total_steps$daily_steps, main = "Histogram of Daily Steps", xlab="Total Number of Steps Taken Each Day")
```
Note the difference in distribution in comparison to the earlier histogram with NAs removed, rather than imputed.

Calculate the mean:
``` {r newmeancalc}
mean(total_steps$daily_steps, na.rm = TRUE)
```
Calculate the median:
``` {r newmediancalc}
median(total_steps$daily_steps, na.rm = TRUE)
```
Both the new mean and median are higher than the old mean and median.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

``` {r weekday}
activity_impute$date <- as.Date(activity_impute$date)
activity_impute$day <- ifelse(weekdays(activity_impute$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
activity_impute$day <- as.factor(activity_impute$day)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

``` {r dualplot}
library(ggplot2)
average_steps <- aggregate(steps ~ interval + day, data=activity_impute, FUN='mean')
g <- ggplot(average_steps, aes(interval, steps)) +
     geom_line() +
     facet_grid(day ~ .) +
     xlab("Interval") + ylab("Number of steps")
g
```

Based on the graph, weekdays have higher peak number of steps in comparison to weekends. However, the peaks during weekends have more even distribution than weekdays.























